\topglue 0.5in
\centerline{\tt Knuth 7}
\vskip 0.5in

\noindent
{\bf Problem 2} We can start with any solution to the normal problem (where we count
from 1 to $n$ instead of 0 to $n-1$ and simply
append 00 on the front.  Therefore, we must have $n-1 = 4m - 1$ or $n-1 = 4m$ --
that is, $n \bmod 4$ is 0 or 1.

\vskip 0.1in
\noindent
{\bf Problem 5} There are $2 n \choose 2$ pairs of positions (since we don't care about 
the ordering of each pair.  Of those, $2 n - k - 1$ satisfy the condition (the first member of the 
pair can be in position 1 up to $2 n - k - 1$ before the second one runs off the end).
A guess as to the number would be to falsely assume the probabilities are independent --
which is clearly nonsense.  The total number of arrangements is $2 n \choose 2, \ldots, 2$,
so the product  ${2 n \choose 2, \ldots, 2} \prod_{j=1}^n \left(2 n - k - 1\right) / {2 n \choose n}$
is an estimate.

\vskip 0.1in
\noindent
{\bf Problem 7} The number of uncompleted pairs $u_i$ must start at 0 at the left of
the sequence (before processing the first number) and end at 0 when we finish.
After each number it can either increase by one or decrease by one.  The longest
sequence we can imagine where we never exceed, 6 is
$\{0, 1, ldots, 5, 6, 5, 6, \ldots, 2, 1, 0\}$.

How can we make use of this information?  Well, one thing we know about Langford
pairs is that each term $k$ contributes a count of 1 to $u$ for $k$ spaces -- that
is, after we process, say, 4, we know that $u$ will be increased by 1 for 4 more spaces
until we close out the pair -- that is, it increases the sum over $u$ by $k+1$.
Thus, $\sum_{i=1}^{2 n} u_i = \sum_{i=1}^{k} k + 1 = {n+1 \choose 2} + n$.  The
other sum, from the above sequence, is, for $u_i$ never exceeding 6,
$11 n - 30$.  So we need $11  n - 30 \ge {n + 1 \choose 2} + n$, or $n \le 15$,
and it doesn't work.

\vskip 0.1in
\noindent
{\bf Problem 39} $G \setminus e$ is always spanning (includes all the vertices of $G$),
but not induced (contains all the edges).

\vskip 0.1in
\noindent
{\bf Problem 40} (a) A spanning subgraph has all $n$ vertices.  So the only
question is which subset of the $e$ edges are present.  There are $2^e$ 
possibilities. (b) Now the question is only which subset of the vertices
are included; there are $2^n$ possibilities (assuming we allow the empty
subgraph).

\vskip 0.1in
\noindent
{\bf Problem 41} (a) 1 and 2. (b) 0 and 3.

\topglue 0.5in
\centerline{\tt Knuth 7.1.1}
\vskip 0.5in

\noindent
{\bf Problem 3} (a) max corresponds to logical or: $\vee$ (b) min corresponds to 
logical and: $\wedge$ (c) $-x$ is negation or right complementation: $\overline{x}$
(d) $x \cdot y$ is equivalence: $x \equiv y$.


\end